---
title: "Data Exploration"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide

---
```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(modelr)
library(haven)
library(plotly)
library(factoextra)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  warning = FALSE,
  message = FALSE
)
```


```{r}
ratings_df = read_csv("data/coffee_ratings.csv") %>% 
  janitor::clean_names() %>%
  select(total_cup_points, species, country_of_origin, 
         region, number_of_bags, bag_weight, grading_date, variety,
         processing_method, aroma, flavor, aftertaste, acidity, body, balance,
         uniformity, clean_cup, sweetness, cupper_points, moisture, color,
         altitude_low_meters, altitude_high_meters, altitude_mean_meters) %>%
  mutate(grading_year = str_extract(grading_date, "\\d{4}"),
         species = as.factor(species),
         variety = as.factor(variety),
         processing_method = as.factor(processing_method),
         moisture = as.factor(moisture),
         color = as.factor(color))
```





### The Distribution of the Total Cup Points

```{r}
points_dis = ggplot(ratings_df, aes(x = total_cup_points)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  coord_cartesian(xlim = c(60, NA)) +
  theme_minimal() +
  labs(title = "Total Cup Points Distribution", x = "Total Cup Points", y = "Frequency")

ggplotly(points_dis)
```











### The Origins of Top Coffees Each Year from 2010 to 2018 + species + variety

```{r}
top_origins_yearly = 
  ratings_df |> 
  group_by(grading_year) |> 
  arrange(grading_year, desc(total_cup_points)) |> 
  slice(1)

top_origins_yearly |> 
  arrange(grading_year) |> 
  select(grading_year, country_of_origin, total_cup_points) |> 
  knitr::kable(
    col.names = c("Year", "Country of Origin", "Total Cup Points"),
    caption = "Top Coffee by Country from 2010 to 2018",
  )
```

The diversity of countries shows that competition in the coffee industry is global, with no single country dominating year after year.

Brazil has been on the list twice (2011 and 2013), indicating that it has consistently produced highly rated coffee during this period.

Ethiopia showed pretty good coffee strength with a top score of 90.58 in 2015, which is the highest coffee rating in those eight years.


### Top 10 Countries with High Coffee Ratings Within the Eight Years

```{r}
top_10_countries = 
  ratings_df |> 
  group_by(country_of_origin) |> 
  summarize(
    avg_total_cup_points = mean(total_cup_points, na.rm = TRUE),
    total_cup_points = sum(total_cup_points, na.rm = TRUE),
    num_entries = n()
  ) |> 
  arrange(desc(avg_total_cup_points)) |> 
  slice_head(n = 10)

top_10_countries |> 
  ggplot(aes(x = reorder(country_of_origin, avg_total_cup_points),
             y = avg_total_cup_points,
             fill = country_of_origin)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(
    title = "Top 10 Countries with High Coffee Ratings 
    (2010-2018)",
    x = "Country",
    y = "Average Total Cup Points",
    fill = "Country"
  ) 
```


### Distribution of Coffee Species

```{r}
ratings_df |> 
  ggplot(aes(x = species)) +
  geom_bar(fill = "orange") +
  labs(
    title = "Distribution of Coffee Species",
    x = "Species",
    y = "Count"
  ) +
  theme_minimal()
```

### Proportion of Processing Methods

```{r}
ratings_df |> 
  count(processing_method) |> 
  mutate(prop = n / sum(n)) |> 
  ggplot(aes(x = reorder(processing_method, prop), 
             y = prop, fill = processing_method)) +
  geom_col() +
  coord_flip() +
  labs(title = "Proportion of Processing Methods", 
       x = "Processing Method", 
       y = "Proportion") +
  theme_minimal()
```

### Total Cup Points by Processing Methods 改成plot_ly 和outlier

```{r}
ratings_df |> 
  drop_na(total_cup_points, processing_method) |> 
  ggplot(aes(x = processing_method, y = total_cup_points, fill = processing_method)) +
  geom_boxplot() +
  labs(title = "Total Cup Points by Processing Methods", 
       x = "Processing Method", 
       y = "Total Cup Points") +
  theme_minimal() +
  coord_flip()
```



### Distribution of Total Cup Points by Species drop na

```{r}
ratings_df |> 
  plot_ly(x = ~ species, y = ~ total_cup_points, type = "box", color = ~species) |> 
   layout(title = "Boxplot of Total Cup Points by Species",
         xaxis = list(title = "Species"),
         yaxis = list(title = "Total Cup Points"))
```


```{r}
yearly_data = ratings_df |> 
  group_by(grading_year) |> 
  summarise(avg_total_cup_points = mean(total_cup_points, na.rm = TRUE)) |> 
  plot_ly(x = ~ grading_year, y = ~ avg_total_cup_points, type = "scatter", 
          mode = "lines+markers") |> 
  layout(title = "Trend of Average Total Cup Points Over Years",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Average Total Cup Points"))
```





```{r}
ratings_df |> 
  filter(country_of_origin %in% c("Ethiopia", "Colombia", "Brazil")) |> 
  ggplot(aes(x = country_of_origin, y = aroma, fill = country_of_origin)) +
  geom_boxplot() +
  labs(title = "Distribution of Aroma Ratings by Country",
       x = "Country", y = "Aroma") +
  theme_minimal()
```






### 我再想想 Differences between different regions of coffee within the same country

```{r}
region_analysis =
  ratings_df |> 
  group_by(country_of_origin, region) |> 
  summarize(
    avg_acidity = mean(acidity, na.rm = TRUE),
    avg_sweetness = mean(sweetness, na.rm = TRUE)
  ) 

region_analysis |> 
  pivot_longer(
    cols = c(avg_acidity, avg_sweetness), 
    names_to = "attribute", 
    values_to = "value") |> 
  ggplot(aes(x = region, y = value, fill = attribute)) +
  geom_boxplot() +
  facet_wrap(~country_of_origin)
```





